Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the `eechidna` `R` package.

### 2019-03-31 draft-paper.Rmd
present tense

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3% and 53.5% (in favour of the Liberal party) over this period. This study explores how electoral socio-demographic characteristics relate to two-party preference, and whether their effects have changed over time. Electoral socio-demographics are derived from the Australian Census, and vote counts are obtained from federal elections.

Joining these two data sources is problematic as there in an inherent asynchronicity in the two types of events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas a federal election (conducted by the Australian Electoral Commission (AEC)) usually occurs every three years. The first problem addressed is that of constructing appropriate Census data for the 2004, 2007, 2010 and 2013 elections - election years in which a Census does not occur. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching electoral data. To obtain appropriate Census data for these four elections, electoral socio-demographics are constructed using a spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in Section \@ref(data). All data and associated documentation relating to this procedure are available in the `eechidna` `R` package [@eechidna], providing a resource for future analysis .

@@

Between 2001 and 2016 there were six elections and four Censuses (see Figure \@ref(fig:timeline)). Electoral boundaries are redistributed regularly by the AEC, meaning that only in the years where both a Census and election occur will the boundaries match - the case for the 2001 and 2016 election. Therefore, for the four elections between 2004 and 2013, both temporal and spatial differences in electorates need to be accounted for when joining the electoral two-party preferred vote with Census data. For these elections a spatio-temporal imputation method is employed to obtain electoral socio-demographics. This method uses Census information from both before and after the election of interest.

@@

ggplot(data=hume_area13) +
  geom_polygon(aes(x=long, y=lat, group=group, fill = elect_div == "HUME"),
               colour="grey50", alpha = 0.4) +
  scale_fill_manual(name="Boundary", values=c("white", "purple"), labels = c("Other 2013 Electorates", "Hume 2013")) +
  theme_map() + coord_equal()
```

@@

ggplot(data=hume_area16) +
  geom_polygon(aes(x=long, y=lat, group=group, fill = year == "2013", 
    alpha = year == "2013", colour=year == "2013")) +
  scale_fill_manual(name="Boundary", values=c("grey95", "purple"), labels = c("2016 Electorates", "Hume 2013")) +
  scale_alpha_manual(values=c(0, 0.4)) +
  scale_color_manual(values=c("grey50", NA)) +
  theme_map() + coord_equal() + guides(alpha = F, color = F)
```

@@

## Influential socio-demographics
To investigate the socio-demographics that have a strong effect on the two-party preferred vote, partial residual plots are used and shown in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2). The partial residuals are the residuals from the fitted model with the estimated effect an individual variable added. The estimated intercept is also added to the partial residuals for interpretability. These show the direction, size and significance of an estimated effect - the slope of the prediction line matches the estimated coefficient, and the shaded region represents a 95% confidence band. If a horizontal line can be drawn through the confidence band, then the effect is insignificant. Plots for each election are faceted to compare the effects over time in Figure \@ref(fig:partresplot1) and Figure \@ref(fig:partresplot2).

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps it is due to these policies, that higher income electorates appear more likely to support the Liberal party, as the `Incomes` factor has a positive effect on Liberal preference (see row 1 in Figure \@ref(fig:partresplot1)). This effect is significant in every election aside from 2004, where it is only marginally insignificant ($p = 0.0613$). Unemployment however, has not been as influential. In 2001 and 2004, electorates with higher unemployment were more likely to support Labor, but over time this has shifted towards support for the Liberal party, culminating in a significantly positive effect in 2016.

@@

### Household mobility
In each of the six elections, electorates with a higher proportion of people that have recently (past five years) moved house were more likely to support the Liberal party, although this effect was marginally insignificant in 2013 (see row 6 in Figure \@ref(fig:partresplot1). Having controlled for characteristics of house ownership and rental prices (via the factors `PropertyOwned` and `RentLoan` respectively), this is effect is somewhat surprising.

>>>>

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3% and 53.5% (in favour of the Liberal party) over this period. This study explores how electoral socio-demographic characteristics relate to two-party preference, and whether their effects have changed over time. Electoral socio-demographics are derived from the Australian Census, and vote counts are obtained from Australian federal elections.

Joining these two data sources is problematic as there in an inherent asynchronicity in the two types of events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas a federal election (conducted by the Australian Electoral Commission (AEC)) usually occurs every three years. The first problem addressed is that of constructing appropriate Census data for the 2004, 2007, 2010 and 2013 elections - election years in which a Census does not occur. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see @DavisStimson98, @Stimson06, @Liao09 and @Stimson09). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching with electoral data. To obtain appropriate Census data for these four elections, electoral socio-demographics are constructed using a spatio-temporal imputation that combines areal interpolation [@Goodchild1993] and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in Section \@ref(data). All data and associated documentation relating to this procedure are available in the `eechidna` `R` package [@eechidna], providing a resource for future analysis .

@@

Between 2001 and 2016 there were six elections and four Censuses (see Figure \@ref(fig:timeline)). Electoral boundaries are redistributed regularly by the AEC, meaning that only in the years where both a Census and election occur will the boundaries match - the case for the 2001 and 2016 elections. Therefore, for the four elections between 2004 and 2013, both temporal and spatial differences in electorates need to be accounted for when joining the electoral two-party preferred vote with Census data. For these elections a spatio-temporal imputation method is employed to obtain electoral socio-demographics. This method uses Census information from both before and after the election of interest.

@@

ggplot(data=hume_area13) +
  geom_polygon(aes(x=long, y=lat, group=group, fill = elect_div == "HUME"),
               colour="grey50", alpha = 0.4) +
  scale_fill_manual(name="Electoral boundaries", values=c("white", "purple"), labels = c("Other 2013 Electorates", "Hume 2013")) +
  theme_map() + coord_equal()
```

@@

ggplot(data=hume_area16) +
  geom_polygon(aes(x=long, y=lat, group=group, fill = year == "2013", 
    alpha = year == "2013", colour=year == "2013")) +
  scale_fill_manual(name="Electoral boundaries", values=c("white", "purple"), labels = c("2016 Electorates", "Hume 2013")) +
  scale_alpha_manual(values=c(0, 0.4)) +
  scale_color_manual(values=c("grey50", NA)) +
  theme_map() + coord_equal() + guides(alpha = F, col = F)
```

@@

## Influential socio-demographics
To investigate the socio-demographics that have a strong effect on the two-party preferred vote, partial residual plots are used and shown in Figures \@ref(fig:partresplot1) and \@ref(fig:partresplot2). The partial residuals are the residuals from the fitted model with the estimated effect an individual variable added. These show the direction, size and significance of an estimated effect - the slope of the prediction line matches the estimated coefficient, and the shaded region represents a 95% confidence band. If a horizontal line can be drawn through the confidence band, then the effect is insignificant. The estimated intercept is also added to the partial residuals for interpretability. Plots for each election are faceted to compare the effects over time in Figure \@ref(fig:partresplot1) and Figure \@ref(fig:partresplot2).

@@

Typically the Labor party campaigns on more progressive policies, which often include tax reform that adversely affects higher income earners, and more generous social assistance programs. Perhaps it is due to these policies, that higher income electorates appear more likely to support the Liberal party, as the `Incomes` factor has a positive effect on Liberal preference (see row 1 in Figure \@ref(fig:partresplot1)). This effect is significant in every election aside from 2004, where it is only marginally insignificant ($p = 0.0613$). Unemployment however, is not as influential. In 2001 and 2004, electorates with higher unemployment align with Labor, but over time this shifts towards support for the Liberal party, culminating in a significantly positive effect in 2016.

@@

### Household mobility
In each of the six elections, electorates with a higher proportion of people that have recently (past five years) moved house are more likely to support the Liberal party, although this effect was marginally insignificant in 2013 (see row 6 in Figure \@ref(fig:partresplot1). Having controlled for characteristics of house ownership and rental prices (via the factors `PropertyOwned` and `RentLoan` respectively), this is effect is somewhat surprising.
